---
title: "lab_02"
author: "derek willis"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

```{r}
new_voters <- md_voters_2024 |> filter(STATE_REGISTRATION_DATE >= '2024-07-21') 
    
```
```{r}
new_voters |>
  group_by(COUNTY) |>
  summarise(
    count_county = n()
  ) |>
  arrange(desc(count_county))
```


**A1** 17,106 new voters (I counted the rows), Montgomery, Baltimore City & Baltimore County saw the highest numbers of registration.



**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2** When looking at the data, more people registered to vote  in the weeks later on in the year, like (28, 30, 31,etc.) which was around the time/leading up to when Biden dropped out of the race. 

```{r}
md_voters_2024 <- md_voters_2024 |>
  mutate(week_number=week(COUNTY_REGISTRATION_DATE))
```
```{r}
md_voters_2024 |>
  group_by(week_number) |>
  summarise(
    total_voter_count=n()
  ) |>
  arrange(desc(total_voter_count))
```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**A3** July saw the largest number of registrations, which makes sense since that was when Biden dropped out.

```{r}
md_voters_2024 <- md_voters_2024 |>
  mutate(month=month(COUNTY_REGISTRATION_DATE))
```
```{r}
md_voters_2024 |>
  group_by(month) |>
  summarise(
    count_month=n()
  ) |>
  arrange(desc(count_month))
```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4** Democratic and unaffiliated voters in the Capital region recorded the two highest numbers of registration. The Central region's democratic and unaffiliated voters came in third and fourth.

```{r}
# ChatGPT recommended I use %in% instead so the counties would be spelled right. 

md_voters_2024 <- md_voters_2024 |> 
  mutate(region = case_when(
    COUNTY %in% c("Allegany", "Garrett", "Washington") ~ "Western",
    COUNTY %in% c("Baltimore", "Baltimore City", "Carroll", "Harford", "Howard") ~ "Central",
    COUNTY %in% c("Montgomery", "Prince George's") ~ "Capital",
    COUNTY %in% c("Anne Arundel", "Calvert", "Charles", "St. Mary's") ~ "Southern",
    COUNTY %in% c("Caroline", "Cecil", "Dorchester", "Kent", "Queen Anne's", "Somerset", "Talbot", "Wicomico", "Worcester") ~ "Eastern Shore",
    )
  )
```
```{r}
md_voters_2024 |>
  group_by(region, PARTY) |>
  summarise(
    total_region=n()
  ) |>
  arrange(desc(total_region))
```


**Q5** What's an additional question you could ask of this data?

**A5** What percentage of new voters are 18-21? 22-25? I'd want to know more about the age demographic with the highest number of new voters.
