---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```



```{r}

#print will take a string and return it in the terminal
print{"Hello world"}

```

```{r}
number <- 2
#the left-facing arrow, called the assignment operator, assigns something to variables
```

Ch. 10 - If you write the formula 'number*number', it will spit out the answer for those variables assigned by the assignment operator

Ch. 13 - CSV data is short for Comma Separated Values data, which is data that has a comma between each column of data.

```{r}

umd_courses <- read_rds("umd_courses.rds")
#the first part - umd_courses - is the name of a variable.
##a variable is just a ma,e to refer to some more complex thing. In this case, that is the data we're importing into R that will be stored as a dataframe, which is one way R stores data.
###variable names are one word all lower case (or two or more words connected by an underscore)
####you can end a variable with a number, but you can't start one with a number

##read_rds() is a function that only works when we've loaded the tidyverse
##a function is a bit of computer code that takes in information and follows a series of pre-determined steps and spits it back out

#inside of the read_rds() function, we've put the name of the file we want to load
##things we put inside of a function to customize what the function does are called arguments

##in R Studio's environment window, the rows are called observations and columns are called variables
```

```{r}
|>
  #this operator is called the pipe operator
  ##keyboard shortcut is cmd-shift-m 
  
%>%
  # this does the same thing as the operator above
  
data |> function

#this basically means, take your data set and do this specific action to it
```

```{r}
n()
#a function that counts the number of rows or records in each group; n is a common symbol used to denote a count of something

arrange()
#it arranges data in order
##by default, it's in ascending order - smallest to largest
```

##Q1 = functions like a notebook
##Q2 = environment
##Q3 = console, where code is going to run
##Q4 = stores our files on our computer

Pre-Lab 2 Notes
------------------
#Filtering creates a subset of the data based on criteria 
##Filtering works with rows -- when we filter, we get fewer rows back than we start with

#Selecting simply returns only the fields named
##Selecting works with columns. You will have the same number of records when you are done, but fewer columns of data to work with

## == equal to
## != not equal to
## > greater than
## >= greater than or equal to
##Be careful: = is not == and = is not "equal to". = is an assignment operator in most languages - how things get named

#AND is &
# OR is |
## OR is additive, AND is restrictive

```{r}
# better ordering?
primary_24 |>
  select(office_name, office_district, candidate_name, party, county_name, election_day, votes) |>
  mutate(
  percent_election_day = (election_day/votes)*100
)  |> 
  arrange(desc(percent_election_day))
```
##the code above arranged the percentage of election day votes to descending order

##`str_to_upper` converts character colums into all uppercase

## If you don't assign the results of a mutate() function to a variable, those changes aren't permanent.

##**Filtering creates a subset of the data based on criteria**. All records where the amount is greater than 150,000. All records that match "College Park". Something like that. **Filtering works with rows -- when we filter, we get fewer rows back than we start with.**

**Selecting simply returns only the fields named**. So if you only want to see city and amount, you select those fields. When you look at your data again, you'll have two columns. If you try to use one of your columns that you had before you used select, you'll get an error. **Selecting works with columns. You will have the same number of records when you are done, but fewer columns of data to work with.**

**Mutate - using data you have to answer a new question

** cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names() |> 
  rename(source = grantor) |> 
  mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description)) |> 
  distinct() |>
  mutate(zip5 = str_sub(zip_code, start=1L, end=5L))


# display the cleaned dataset
cleaned_md_grants_loans
##allows us to clean up columns stored as character strings

**the rename function allows us to remove that x in front of line number column
##cleaned_conowingo <- conowingo |>
  clean_names() |> 
  rename(linenumber = x1_linenumber)

# display the cleaned dataset
cleaned_conowingo

**gets rid of duplicates in datasets
# cleaning function
cleaned_conowingo <- conowingo |>
  clean_names() |> 
  rename(linenumber = x1_linenumber) |> 
  mutate(amount = as.numeric(amount)) |> 
  distinct()
  

# display the cleaned dataset
cleaned_conowingo
