---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```



```{r}

#print will take a string and return it in the terminal
print{"Hello world"}

```

```{r}
number <- 2
#the left-facing arrow, called the assignment operator, assigns something to variables
```

Ch. 10 - If you write the formula 'number*number', it will spit out the answer for those variables assigned by the assignment operator

Ch. 13 - CSV data is short for Comma Separated Values data, which is data that has a comma between each column of data.

```{r}

umd_courses <- read_rds("umd_courses.rds")
#the first part - umd_courses - is the name of a variable.
##a variable is just a ma,e to refer to some more complex thing. In this case, that is the data we're importing into R that will be stored as a dataframe, which is one way R stores data.
###variable names are one word all lower case (or two or more words connected by an underscore)
####you can end a variable with a number, but you can't start one with a number

##read_rds() is a function that only works when we've loaded the tidyverse
##a function is a bit of computer code that takes in information and follows a series of pre-determined steps and spits it back out

#inside of the read_rds() function, we've put the name of the file we want to load
##things we put inside of a function to customize what the function does are called arguments

##in R Studio's environment window, the rows are called observations and columns are called variables
```

```{r}
|>
  #this operator is called the pipe operator
  ##keyboard shortcut is cmd-shift-m 
  
%>%
  # this does the same thing as the operator above
  
data |> function

#this basically means, take your data set and do this specific action to it
```

```{r}
n()
#a function that counts the number of rows or records in each group; n is a common symbol used to denote a count of something

arrange()
#it arranges data in order
##by default, it's in ascending order - smallest to largest
```

##Q1 = functions like a notebook
##Q2 = environment
##Q3 = console, where code is going to run
##Q4 = stores our files on our computer